The core transcriptome of mammalian placentas and the divergence of expression with placental shape

### 2016-11-17 manuscript/manuscript.Rnw
move human specific up
discuss EGFR signalling

\begin{sloppypar}
The quality of reads was evaluated using FASTQC v0.11.2
(\url{http://www.bioinformatics.babraham.ac.uk/projects/fastqc/}).
Transcriptomes without a suitable reference genome (\smonkey, \galili,
\carmeli; italic reference in Supplemental
\cref{tab:ref_genome_aligned_species}) were assembled using Trinity
v2014-07-17 \cite{Haas.ea2013:Denovotranscriptsequence} and annotated
using Diamond \cite{Buchfink.ea2015:Fastandsensitiveprotein} against
either the \human or \mouse transcriptome (Supplemental
\cref{tab:ref_genome_aligned_species}). For annotation using Diamond,
the match with the highest bitscore was chosen as the annotation;
transcripts where the highest scoring match had a bitscore below 40
were left unannotated.
% 
All transcriptomes were then aligned to the appropriate reference (or
assembled) genome using STAR v2.4.2a
\cite{Dobin.ea2013:STARultrafastuniversalRNA-seq} and expression
values in \ac{FPKM}
\cite{Trapnell.ea2010:Transcriptassemblyandquantification} were
quantified using Cufflinks v2.2.1
\cite{Trapnell.ea2009:TopHatdiscoveringsplicejunctions} (Supplemental
\cref{tab:all_ateles_fusciceps_placenta,tab:all_homo_sapiens_placenta,tab:all_pan_paniscus_placenta,tab:all_mus_musculus_placenta,tab:all_spalax_galili_placenta,tab:all_spalax_carmeli_placenta,tab:all_bos_taurus_placenta,tab:all_ovis_aries_placenta,tab:all_sus_scrofa_placenta,tab:all_equus_caballus_placenta,tab:all_canis_familiaris_placenta,tab:all_loxodonta_africana_placenta,tab:all_dasypus_novemcinctus_placenta,tab:all_monodelphis_domestica_placenta}).
%
The genome of the closely-related \textit{P. troglodytes} was used as
a reference for \bonobo. The separation between the two recognized Pan
species is generally perceived to be $\approx 1$ million years
\cite{Prüfer.ea2012:bonobogenomecomparedwith}. %
%
\end{sloppypar}

@@

All 1:1 orthologs present among \opossum, \textit{Pan troglodytes},
\cow, \dog, \elephant, \mouse, and \human were determined using
Ensembl v.80 gene trees via Biomart
\cite{Kasprzyk2011:BioMartdrivingparadigmchange} using the biomaRt R
package
\cite{Durinck.ea2009:Mappingidentifiersforintegration,Durinck.ea2005:BioMartandBioconductorpowerful}.
A total of \Sexpr{combined.fpkm[species!="homo sapiens" & mean_fpkm >=
  1,length(unique(human_name))]} 1:1 orthologous genes were expressed
in at least one species. Of these, \Sexpr{combined.fpkm[species!="homo
  sapiens" & mean_fpkm >= 100,length(unique(human_name))]} 1:1
orthologous genes with \ac{FPKM} $\ge 100$ were retained for future
analyses of gene expression evolution described below.

@@

The core placenta transcriptome is significantly enriched in genes
which are involved in multiple biological pathways, including multiple
pathways related to translation (ribosome, translation, signal
recognition particle targeting to the rough endoplasamic reticulum,
etc.) (Supplemental \cref{tab:core_placenta_pathway_analysis}). Beyond
translation, the next most abundant set of significantly enriched
molecular pathways are related to the migration of the trophoblast
into the endometrium, including the EGFR1 pathway and the Actin Y
pathway. Two other significant pathways, the mPR/Progesterone pathway
and Rho cell motility pathways, feed into the Actin Y pathway as well.
Supplemental \cref{fig:egf_kegg_pathway} shows the EGF pathway in
humans with the median expression in FPKM of genes in the pathway
shown. Notably, many of the genes on the GRB2→MYC pathway are strongly
expressed; their activation leads to an increase in migration and
invasion in response to EGF binding to
EGFR1~\cite{Gupta.ea2016:CellSignalingPathwaysInvolved}. The Actin
cytoskeleton pathway is depicted in Supplemental
\cref{fig:kegg_actin}, which shows that \textit{FN1} and \textit{ITG}
are highly expressed, which activate Rac, and potentially lead to the
formation of focal adhesions and lamellipodia formation through their
inhibition of GSN and HSPC300. Other compoenents of cell-cell
adhesion, including \textit{F11R} which forms tight junctions in
epithelia, proteglycans \textit{SDC4} and \textit{LUM}, and the
laminin \textit{LAMC1}.

@@

\subsubsection{Immune Regulation and Annexins}
The transcriptomes from multiple different species enabled us to
identify the set of expressed genes which are common to all placentas.
Components of the annexin complexes, including
\textit{ANXA2}, \textit{ANXA1}, \textit{S100A11}, and \textit{S100A10}
are expressed in all placentas examined, and argue for the evolution
importance of annexins in placenta function, where they likely
establish maternal-fetal
tolerance\cite{Liu.ea2015:AnnexinA2complexeswith,Hutchinson.ea2011:Molecularregulatorsofresolution,Aarli.ea1997:Suppressiveeffectonlymphoproliferation}.
We suggest that further research into molecular role of these annexin
complexes may illuminate how immune tolerance is maintained during
placentation. The core placenta transcriptome is also enriched in
genes involved in maintaining the structure of the placenta and
transporting proteins, as would be expected given the function of the
placenta.

@@

The high level of expression of components of the EGFR signaling
pathway in the core placenta transcriptome (including \textit{GRB2}, \textit{EGFR}, \textit{MAPK3}, \textit{MAP2K2}, \textit{SHC1}, \textit{AKT3}, \textit{AKT2}, \textit{ERBB3}

% Role of CD9 in invasion; commonality between placentas

>>>>

\begin{sloppypar}
The quality of reads was evaluated using FASTQC v0.11.2
(\url{http://www.bioinformatics.babraham.ac.uk/projects/fastqc/}).
Transcriptomes without a suitable reference genome (\smonkey, \galili,
\carmeli; italic reference in Supplemental
\cref{tab:ref_genome_aligned_species}) were assembled using Trinity
v2014-07-17 \cite{Haas.ea2013:Denovotranscriptsequence} and annotated
using Diamond \cite{Buchfink.ea2015:Fastandsensitiveprotein} against
either the \human or \mouse transcriptome (Supplemental
\cref{tab:ref_genome_aligned_species}). For annotation using Diamond,
the match with the highest bitscore was chosen as the annotation;
transcripts where the highest scoring match had a bitscore below 40
were left unannotated.
% 
All transcriptomes were then aligned to the appropriate reference (or
assembled) genome using STAR v2.4.2a
\cite{Dobin.ea2013:STARultrafastuniversalRNA-seq} and expression
values in \ac{FPKM}
% \cite{Trapnell.ea2010:Transcriptassemblyandquantification}
were
quantified using Cufflinks v2.2.1
\cite{Trapnell.ea2009:TopHatdiscoveringsplicejunctions} (Supplemental
\cref{tab:all_ateles_fusciceps_placenta,tab:all_homo_sapiens_placenta,tab:all_pan_paniscus_placenta,tab:all_mus_musculus_placenta,tab:all_spalax_galili_placenta,tab:all_spalax_carmeli_placenta,tab:all_bos_taurus_placenta,tab:all_ovis_aries_placenta,tab:all_sus_scrofa_placenta,tab:all_equus_caballus_placenta,tab:all_canis_familiaris_placenta,tab:all_loxodonta_africana_placenta,tab:all_dasypus_novemcinctus_placenta,tab:all_monodelphis_domestica_placenta}).
%
The genome of the closely-related \textit{P. troglodytes} was used as
a reference for \bonobo. The separation between the two recognized Pan
species is generally perceived to be $\approx 1$ million years
\cite{Prüfer.ea2012:bonobogenomecomparedwith}. %
%
\end{sloppypar}

@@

All 1:1 orthologs present among \opossum, \textit{Pan troglodytes},
\cow, \dog, \elephant, \mouse, and \human were determined using
Ensembl v.80 gene trees via Biomart using the biomaRt R package
\cite{Durinck.ea2009:Mappingidentifiersforintegration}. %
A total of \Sexpr{combined.fpkm[species!="homo sapiens" & mean_fpkm >=
  1,length(unique(human_name))]} 1:1 orthologous genes were expressed
in at least one species. Of these, \Sexpr{combined.fpkm[species!="homo
  sapiens" & mean_fpkm >= 100,length(unique(human_name))]} 1:1
orthologous genes with \ac{FPKM} $\ge 100$ were retained for future
analyses of gene expression evolution described below.

@@

The core placenta transcriptome is significantly enriched in genes
which are involved in multiple biological pathways, including multiple
pathways related to translation (ribosome, translation, signal
recognition particle targeting to the rough endoplasamic reticulum,
etc.) (Supplemental \cref{tab:core_placenta_pathway_analysis}). Beyond
translation, the next most abundant set of significantly enriched
molecular pathways are related to the migration of the trophoblast
into the endometrium, including the EGFR1 pathway and the Actin Y
pathway. Two other significant pathways, the mPR/Progesterone pathway
and Rho cell motility pathways, feed into the Actin Y pathway as well.
Supplemental \cref{fig:egf_kegg_pathway} shows the EGF pathway in
humans with the median expression in FPKM of genes in the pathway
shown. Notably, many of the genes on the GRB2→MYC pathway are strongly
expressed; their activation leads to an increase in migration and
invasion in response to EGF binding to
EGFR1~\cite{Gupta.ea2016:CellSignalingPathwaysInvolved}. The Actin
cytoskeleton pathway is depicted in Supplemental
\cref{fig:kegg_actin}, which shows that \textit{FN1} and \textit{ITG}
are highly expressed, which activate Rac, and potentially lead to the
formation of focal adhesions and lamellipodia formation through their
inhibition of GSN and HSPC300. Other components of cell-cell
adhesion, including \textit{F11R} which forms tight junctions in
epithelia, proteglycans \textit{SDC4} and \textit{LUM}, and the
laminin \textit{LAMC1}.

@@

\subsubsection{Immune Regulation and Annexins}
The transcriptomes from multiple different species enabled us to
identify the set of expressed genes which are common to all placentas.
Components of the annexin complexes, including
\textit{ANXA2}, \textit{ANXA1}, \textit{S100A11}, and \textit{S100A10}
are expressed in all placentas examined, and argue for the evolution
importance of annexins in placenta function, where they likely
establish maternal-fetal
tolerance\cite{Liu.ea2015:AnnexinA2complexeswith,Hutchinson.ea2011:Molecularregulatorsofresolution,Aarli.ea1997:Suppressiveeffectonlymphoproliferation}.
We suggest that further research into molecular role of these annexin
complexes may illuminate how immune tolerance is maintained during
placentation.

@@

The high expression of components of the EGFR signaling pathway in the
core placenta transcriptome (including \textit{GRB2}, \textit{RALB})
as well as annexins which are involved in the location of EGF receptor
localization (\textit{ANXA1} and \textit{ANXA2}) in the core placenta
is consistent with previous findingsshowing that alterations to the
EGF signaling system change invasion of trophoblast cells into the
spiral arteries~\cite{Armant.ea2015:Reducedexpressionofepidermal}.

% TAGLN2 -- likely to be produced by the placenta as well; possibly
% due to the smooth muscle transition?
